Lossy compression of DNA sequencing quality data

### 2015-05-02 manuscript/q-compression.tex
More polishing text and making the document more concise.

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool that models quality values as a multivariate
Gaussian distribution, computing the mean and covariance for each
position in the read\citep{Ochoa:2013rt}. Once the model parameters
are calculated, they are stored by the decoder to later reconstruct a
\emph{representative} quality value.  \textsc{QualComp} takes as input
a user-specified rate (bits/read) and then poses an optimization
problem of how to allot these bits for a given position within the
read while minimizing the \textsc{mse}. The quality values can be
clustered before hand to produce more accurate models.

@@

\subsection{Lossy compression acceptable for subsequent biological analyses}

The primary concern of using lossy compression methods is naturally
the extent of information loss, that we quantified by \textsc{mse} in
this study. \textsc{mse} and compressibility provide information in
the theoretical context to the methods, but they are not the end-all
of evaluation criteria. The performance of compressed datasets in
different subsequent analyses and applications are just as
important. Our benchmarks showed that some of the compression methods
with high error rates are still practical for certain kinds of
applications. Many subsequent tools proved to have enough additional
redundancy built-in to handle such loss in information. Passing the
decompressed quality values through quality control software shows
that most methods filter nearly as many bases as using original
quality sequences. Assemblers performing sequence alignment use
percent similarity scores that are typically robust to standard
sequencing errors.

It is important to note that the downstream applications dictate
which lossy compression methods are acceptable. As noted
above, 2-bin encoding is able to retain information about low
quality bases that occur within stretches of high quality bases,
whereas the polynomial regression methods tend to smooth these
qualities. In addition to aligning less reads that the 2-bin
encoding, the polynomial regression methods may also be hindering
downstream tools that use the alignment results. A \textsc{snp}
detection pipeline will commonly use the quality values to record if a
potential difference from the reference is the result of genetic
variation or sequencing error.  Erroneously marking a base as higher
quality, may lead to the base being incorrectly labeled as a
\textsc{snp}.  2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly. In the case of assembly,
the 7th-degree polynomial regression method outperforms 2-bin encoding
in nearly all metrics.

\subsection{Extension of 2-bin encoding}

2-bin encoding has the nice property of being simple to compute and
has good bits/base-call values. The 2-bin encoding suffers from having
a high \textsc{mse}, but fortunately, we have shown that in the case
of genome assembly, 2-bin encoding outperforms all polynomial
regressions encodings with degree less than 3. 2-bin encoding of the
fragment and short-jump libraries of \textit{Rhodobacter sphaeroides}
have \textsc{mse}s of $2.42\times$ and $10.76\times$ the 3rd-degree
polynomial regression encodings, respectively. This further highlights
the importance of using additional contextual information of the
subsequent analyses when evaluating compressed quality values.

A potential extension to 2-bin encoding is to incorporate an
additional bin (\emph{okay}). The \emph{okay} value can be used where
the base qualities fall within a 2-bin range. Because the distribution
of quality values is skewed towards higher quality, we need to
experiment with different cutoffs for the \emph{okay} value and
determine if the additional storage costs are associated with positive
effects on subsequent analyses.

\subsection{Extension of polynomial regression}

The downside of modeling quality sequences using polynomial regression
is that the model often requires a high number of degrees to achieve
the same \textsc{mse} as the profile and \textsc{QualComp} methods.
However, storing a high number of coefficients requires more storage
than losslessly compressing the original data. In order to increase
the compressibility of modeling, we can attempt to store the
\emph{profiles} of certain polynomial functions. In other words, we
can use a spline (a function that is piecewise-defined by polynomial
functions) to represent a given quality sequence. Similar to our
profile-encoding method, the user can specify how many polynomial
functions they wish to store. Then a quality sequence can be divided
evenly into a given number of segments, and each segment can be
annotated with a polynomial function profile that closely matches its
quality sequence.

>>>>

For comparison to other developed methods we use \textsc{QualComp}, a
lossy compression tool~\citep{Ochoa:2013rt}. The program models
quality values as a multivariate Gaussian distribution, computing the
mean and covariance for each position in the read, and these varlues
are stored by the decoder to later reconstruct a \emph{representative}
quality value. \textsc{QualComp} takes as input a user-specified rate
(bits/read), and these bits are allotted among positions within the
read to minimize the \textsc{mse}. The quality values can be clustered
before hand to produce more accurate models.

@@

\subsection{Lossy compression is acceptable for some subsequent  
analyses}

It is common to evaluate lossy compression methods in terms of
information loss, and some measure of compression effectiveness.
However, from a pragmatic perspective, it is the effects on the
results of subsequent analyses that are most important. Our results
showed that some of the compression methods with relatively high
information lossy (\textsc{mse}) are still practical for some
applications. Many bioinformatics tools proved to be robust to
moderate information loss. Passing the decompressed quality values
through quality control software shows that most methods filter nearly
as many bases as using original quality sequences. Assemblers
performing sequence alignment use percent similarity scores that are
typically robust to standard sequencing errors.

Downstream applications dictate which lossy compression methods are
acceptable. As noted above, 2-bin encoding is able to retain
information about low quality bases that occur within stretches of
high quality bases, whereas the polynomial regression methods tend to
smooth these qualities. In addition to aligning less reads that the
2-bin encoding, the polynomial regression methods may also be
hindering downstream tools that use the alignment results. A
\textsc{snp} detection pipeline will commonly use the quality values
to record if a potential difference from the reference is the result
of genetic variation or sequencing error. Erroneously marking a base
as higher quality, may lead to the base being incorrectly labeled as a
\textsc{snp}.  2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly, and 7th-degree polynomial
regression modeling outperforms it in nearly all metrics.
